---
title: "Reliability and measurement error in psychometrics"
subtitle: "An applied example from Rasch Measurement Theory"
title-block-banner: "#009ca6"
title-block-banner-color: "#FFFFFF"
author: 
  name: Magnus Johansson
  affiliation: RISE Research Institutes of Sweden
  affiliation-url: https://www.ri.se/shic
  orcid: 0000-0003-1669-592X
date: last-modified
date-format: iso
format: 
  html:
    toc: true
    toc-depth: 3
    toc-title: "Table of contents"
    embed-resources: true
    standalone: true
    page-layout: full
    mainfont: 'Lato'
    monofont: 'Roboto Mono'
    code-overflow: wrap
    code-fold: show
    code-tools: true
    code-link: true
    number-sections: true
    fig-dpi: 96
    fig-cap-location: top
    layout-align: left
    linestretch: 1.6
    theme: [materia, custom.scss]
    css: styles.css
    license: CC BY
    email-obfuscation: javascript
  pdf:
    papersize: a4
    documentclass: report 
execute:
  echo: true
  warning: false
  message: false
  cache: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
bibliography: 
  - references.bib
  - grateful-refs.bib
---

## Introduction

::: callout-note
**Purposes of writing this text:** 

- to provide an applied example of how to estimate
latent/factor scores (thetas) and their Standard Error of Measurement
(SEM). 
- demonstrate how SEM relates to the true score, based on
simulation.
- test the functions in [`brms`](https://cran.r-project.org/web/packages/brms/index.html) that enables one to specify
measurement error in a regression model.
:::

This post is mainly about the Standard Error of Measurement (SEM) metric
estimated by Item Response Theory (IRT)/Rasch theta estimation tools, but since this will
probably not be familiar to most readers I will to provide some
additional information and context.

I write this post primarily from the perspective of psychological
measurement using questionnaires with multiple items to assess a latent
variable, such as a depression or well-being questionnaire.

First, I need to point out that the Rasch measurement model allows one
to estimate reliability properties of the set of items themselves,
independent of a sample. This may seem like a strange thing to many
readers, especially if you mostly have experience with psychometrics
within the Classical Test Theory (CTT) paradigm (factor analysis and
other related methods).

Second, the estimated SEM in Rasch/IRT varies across the latent
continuum, it is not a constant point estimate, nor is it the same for
every individual [@samejima1994]. The Fisher information function is
used to estimate a standard error dependent on location on the latent
continuum [@kreiner2012]. This is often described in a curve showing the Test Information Function (see @fig-tifexample). The connection between item threshold locations and TIF can also be seen by reviewing the targeting figure (@fig-targeting) and its middle section that aggregates the threshold locations.

To summarise, here are two things regarding CTT/CFA methods that are not
unknown but very seldom discussed:

-   reliability within CTT is sample dependent. You have most likely
    never read a CTT psychometrics paper that provides information about
    the reliability of the measure or test itself. 
      - This is due to the
    inability of the measurement model to separate person and item
    properties (see ["specific objectivity"](https://scholar.google.com/scholar?hl=en&as_sdt=0%2C5&q=%22specific+objectivity%22)). 
      - This is also a problem with IRT models with more than one parameter[^1].
-   reliability within CTT is always[^2] a constant value claimed to be
    the same for all participants at all levels of the latent variable.
    (I think this is an unreasonable assumption.)
    
There has been a lot of discussion within CTT about
reliability metrics during the last 15 years or so, mostly justified
critique of Cronbach’s alpha [i.e @sijtsma2008; @mcneish2018;
@flora2020; @cortina2020]. But also mostly proposing other sample
dependent metrics that provide a point estimate of reliability (some
argue for also presenting some form of confidence interval).

This is not to say that we should not talk about sample reliability, we absolutely should! But it is important to differentiate between test/measure properties and sample properties. This is particularly important in the context of psychometric analyses and the "validation" of measures, where the intent usually is for others to be able to make use of a measure.

If you are interested in learning more about psychometrics and IRT/Rasch measurement models, you might find these resources helpful:

- [Intro slides](https://pgmj.github.io/RaschIRTlecture/slides.html) that are rather verbose and should help clarify the basic concepts.
- [Our preprint](https://doi.org/10.31219/osf.io/3htzc) on psychometric criteria, which also includes a Rasch psychometric analysis reporting checklist.
- The [package vignette](https://pgmj.github.io/raschrvignette/RaschRvign.html) for the [`RISEkbmRasch`](https://github.com/pgmj/RISEkbmRasch) R package.

[^1]: This is because the discrimination parameter (2PL, GRM, etc) is also sample
    specific/dependent. See Stemler & Naples [-@stemler2021], figure 8,
    for a nice explanation and case example. One exception exists for the 3PL model: if the pseudo-guessing parameter is constant across all items, sample
    independence is upheld [@jiao2023].

[^2]: Except CME/BME methods [@kroc2020], but I have never seen those
    applied or reported in psychometric CTT papers.

## Setup

Many of the functions used in this post are available in my R package for Rasch analysis, [`RISEkbmRasch`](https://github.com/pgmj/RISEkbmRasch), including the function that simulates response data. The package is not available on CRAN, see below (or the link) for installation instructions.

```{r}
# You need to use devtools or remotes to install the package from GitHub.
# First install devtools by:
# install.packages('devtools')
# then 
# devtools::install_github("pgmj/RISEkbmRasch", dependencies = TRUE)
library(RISEkbmRasch) # this also loads a bunch of other packages
library(tidyverse)
library(eRm)
library(catR)
library(readxl)
library(janitor)
library(tinytable)
library(faux)
library(ggrain)
library(patchwork)
library(lme4)
library(brms)
library(modelsummary)
library(broom.mixed)

### some commands exist in multiple packages, here we define preferred ones that are frequently used
select <- dplyr::select
count <- dplyr::count
rename <- dplyr::rename

# get theming/colors
source("RISE_theme.R")
extrafont::loadfonts(quiet = TRUE)
```

We'll use data and item parameters for the Perceived Stress Scale's
seven negative items [@rozental2023].

```{r}
df.all <- read_excel("data/Swedish_PSS_Rasch_analysis.xlsx") %>% 
  select(starts_with("PSS"))

names(df.all) <- paste0("q",c(1:14))

final_items <- c("q1","q2","q3","q8","q11","q12","q14")

df <- df.all %>% 
  select(all_of(final_items))

```

## Ordinal/interval transformation table

```{r}
#| label: tbl-ordint
#| tbl-cap: Ordinal/interval transformation table
RIscoreSE(df, score_range = c(-5,5))
```

This provides us with a lookup table (@tbl-ordint) for transforming ordinal sum scores
from the seven items to logit scores, with their respective SEM. Multiplying SEM by
1.96 gives us a 95% confidence interval. Logit scores and SEM are
estimated using the unbiased Weighted Likelihood method [@warm1989].

The lookup table works for respondents with complete response data. If
you have respondents with missing responses the estimation method we
will use below will take this into account when estimating the SEM. No
need for imputation.

```{r}
#| label: fig-ordint
#| fig-cap: 'Ordinal sum scores and corresponding interval scores with 95% CI'
RIscoreSE(df, score_range = c(-5,5), output = "figure", error_multiplier = 1.96)
```

@fig-ordint illustrates the transformation table
information. We can also look at the SEM separately, across the latent
variable continuum.

```{r}
#| label: fig-sem
#| fig-cap: 'Measurement error across the latent continuum'
ord_int_table<- RIscoreSE(df, 
                          score_range = c(-5,5), 
                          output = "dataframe") %>% 
  janitor::clean_names() 
  
ord_int_table %>% 
  ggplot(aes(x = logit_score, y = logit_std_error)) +
  geom_point() +
  geom_line() +
  coord_cartesian(ylim = c(0,1.6), xlim = c(-5,5)) +
  scale_x_continuous(breaks = seq(-5,5,1)) +
  labs(x = "Interval (logit) score",
       y = "Measurement error")
```

## Simulation

We'll now simulate a dataset with 1000 respondents based on the item
parameters and a vector of latent/factor scores (thetas). Then we'll
estimate thetas and SEM based on the response data generated. Finally,
we'll look at the coverage of estimated values compared to the input
vector of scores.

::: callout-note
This is just a single simulated dataset for illustration, just as the regression example below also only uses one dataset. This is not a simulation study meant to show generalizable results. As such, the results should be interpreted with care. However, the code provided could be helpful if you want to conduct such as study.
:::

### Input parameters

The item parameters need to be in a matrix object, with items as rows
and their thresholds (which may vary in numbers) as columns.

```{r}
# read item parameters file
item_params <- read_csv("data/itemParameters.csv") %>% 
  as.matrix()

item_params
```

### Targeting

Just to get a picture of how the item category thresholds are distributed, we can use the data from our paper to produce a targeting figure.

```{r}
#| label: fig-targeting
#| fig-cap: 'Targeting properties of the PSS-7'

RItargeting(df)
```


### Simulating data

Now we can use the item parameters to simulate data. We'll use `rnorm()`
with a mean that matches the item parameters and SD = 1.5, and generate
2000 theta values. Then these are used to simulate response data for the
7 items.

```{r}
set.seed(1437)

# simulation function needs item parameters as a list object
itemlist <- list(
  i1 = list(item_params[1,]),
  i2 = list(item_params[2,]),
  i3 = list(item_params[3,]),
  i4 = list(item_params[4,]),
  i5 = list(item_params[5,]),
  i6 = list(item_params[6,]),
  i7 = list(item_params[7,])
)

# randomly generated normal distribution of thetas
nsim = 2000
inputThetas <- rnorm(n = nsim, 
                     mean = mean(item_params), # 0.57 
                     sd = 1.5)

# simulate response data based on thetas and items above
simData <- SimPartialScore(
  deltaslist = itemlist,
  thetavec = inputThetas
) %>%
  as.data.frame()
```

### Estimating theta & SEM

Next, we estimate theta values and their SEM from the response data,
using the known item parameters. This is how you would/should use a
calibrated item set with real data as well.

```{r}
# estimate theta values.
est_thetas <- RIestThetas(simData, 
                          itemParams = item_params, 
                          theta_range = c(-5,5)
                          )
# estimate SEM
est_sem <- 
  map_vec(est_thetas, ~ semTheta(thEst = .x,
                                 it = item_params,
                                 model = "PCM",
                                 method = "WL",
                                 range = c(-5, 5)))
                   
```

## Results

```{r}
# get all variables in the same dataframe
sim_data <- simData %>% 
  add_column(est_thetas = est_thetas,
             est_sem = est_sem,
             input_thetas = inputThetas)

```

### Input vs estimated thetas

```{r}
#| label: fig-simestt
#| fig-cap: Comparing input and estimated theta values
sim_data %>% 
  add_column(id = c(1:nsim)) %>% 
  pivot_longer(c(input_thetas,est_thetas)) %>% 
  ggplot(aes(x = name, y = value, fill = name)) +
  geom_rain(id.long.var = "id", alpha = 0.7) +
  scale_y_continuous(limits = c(-5,5), breaks = seq(-5,5,1)) +
  guides(fill = "none") +
  theme_rise() +
  labs(x = "", y = "Theta (logit scale)")
```

As you can see, the generated sample of "input_thetas" has a continuous gaussian distribution, while the estimated thetas are limited to certain values. This is due the the number and distribution of items and item category thresholds (see @fig-targeting).

### Coverage

```{r}
#| label: tbl-coverage
#| tbl-cap: Expected and actual coverage in sample
cov_data <- sim_data %>% 
  mutate(absdiff = abs(est_thetas - input_thetas)) %>%
  mutate(coverage_95 = factor(ifelse(absdiff < 1.96*est_sem, 1, 0)),
         coverage_90 = factor(ifelse(absdiff < 1.645*est_sem, 1, 0)),
         coverage_75 = factor(ifelse(absdiff < 1.15*est_sem, 1, 0))) %>% 
  pivot_longer(cols = c(coverage_95, coverage_90, coverage_75),
               names_to = "coverage",
               values_to = "coverage_n") %>% 
  count(coverage_n, by = coverage) %>% 
  filter(coverage_n == 1) %>% 
  mutate(`Actual coverage` = n/nsim*100, .before = "n") %>% 
  select(!c(coverage_n,n,by)) %>% 
  add_column(`Expected coverage` = c("75%","90%","95%"))

uncov <- 100-cov_data[3,1] %>% pull()
tt(cov_data)
```

```{r}
#| label: fig-estsem
#| fig-cap: Distribution of measurement error across the estimated theta values
sim_data %>% 
  mutate(error = est_thetas - input_thetas,
         absdiff = abs(est_thetas - input_thetas),
         coverage_95 = factor(ifelse(absdiff < 1.96*est_sem, 1, 0))
         ) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  left_join(.,ord_int_table %>% mutate_if(is.numeric, round, 2), by = join_by("est_thetas" == "logit_score")) %>% 
  ggplot(aes(x = est_thetas, y = error)) +
  geom_errorbar(aes(ymin = 0 - (1.96 * logit_std_error),
                    ymax = 0 + (1.96 * logit_std_error)),
                width = 0.1, color = "darkgrey", alpha = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(aes(color = coverage_95), alpha = 0.6) +
  labs(title = "",
       x = "Estimated theta values",
       y = "Measurement error") +
  guides(color = "none") +
  labs(caption = paste0("Red dots (",round(uncov,2),"%) indicate estimated thetas outside of 95% confidence interval (SEM * 1.96, indicated by grey error bars).")) +
  theme_rise()

```

### SEM and Mean Absolute Error (MAE)

```{r}
sim_data %>% 
  mutate(absdiff = abs(est_thetas - input_thetas)) %>% 
  pivot_longer(cols = c(absdiff,est_sem)) %>% 
  mutate(name = dplyr::recode(name, "est_sem" = "SEM",
                       "absdiff" = "MAE")) %>% 
  
  ggplot(aes(x = est_thetas, y = value, color = name, group = name)) +
  geom_point() +
  guides(fill = "none") +
  labs(x = "Estimated theta value",
       y = "Error")
```


```{r}
sim_data %>% 
  mutate(absdiff = abs(est_thetas - input_thetas)) %>% 
  summarise(mae = mean(absdiff))
```

```{r}
mean(sim_data$est_sem)
```

```{r}
#| label: fig-tifexample
#| fig-cap: Test Information Function for simulated response data
#| fig-width: 8
sim_data %>% 
  select(starts_with("i")) %>% 
  select(!input_thetas) %>% 
  RItif(samplePSI = TRUE)
```

## Regression with measurement error

This is just exploratory, and I am no expert in Bayesian models, so please [let me know](mailto:magnus.p.johansson@ri.se) if something should be modified or if there are other models/tools that should be considered.

There is, as far as I know, no implementation of a frequentist regression tool in R that allows one to specify a vector of measurement errors. I have been looking into this recently and found some interesting R packages (`galamm`, `mecor`, and `simex`), but none seem to be (easily) applicable for this type of data and setup.

We'll simulate new data with two "time points", n = 200 each, that have mean
theta locations 1 logit apart, SD = 1.5 for both groups, and ICC = 0.5. Then we'll run a mixed model. Since we only have two time points, we'll use random intercepts for id.

```{r}
# randomly generated normal distributions of thetas with correlation
data <- rnorm_multi(n = 200, 
                    mu = c(-0.5, +0.5),
                    sd = c(1.5, 1.5),
                    r = 0.5, 
                    varnames = c("pre", "post"),
                    empirical = FALSE)

# light wrangling to get long format and an id variable
data_long <- data %>% 
  add_column(id = 1:(nrow(.))) %>% 
  pivot_longer(cols = c("pre", "post"), 
               names_to = "time", 
               values_to = "input_theta") %>% 
  mutate(time = dplyr::recode(time, "pre" = 0, "post" = 1))

# simulate response data based on thetas and items above
sim_data_g <- SimPartialScore(
  deltaslist = itemlist,
  thetavec = data_long$input_theta
) %>%
  as.data.frame()

# estimate theta values.
sim_data_g$est_theta <- RIestThetas(sim_data_g, 
                                    itemParams = item_params, 
                                    theta_range = c(-5,5))
# estimate SEM
sim_data_g$est_sem <- 
  map_vec(sim_data_g$est_theta, ~ semTheta(thEst = .x,
                                           it = item_params,
                                           model = "PCM",
                                           method = "WL",
                                           range = c(-5, 5)))

sim_data_g <- cbind(sim_data_g,data_long)
```


```{r}
sim_data_g %>% 
  group_by(time) %>% 
  summarise(mean_input = mean(input_theta),
            mean_estimated = mean(est_theta),
            median_input = median(input_theta),
            median_estimated = median(est_theta),
            sd_input = sd(input_theta),
            sd_estimated = sd(est_theta),
            mad_input = mad(input_theta),
            mad_estimated = mad(est_theta))
```

### Plot your data

```{r}
#| label: fig-rain
#| fig-cap: Comparison of simulated and estimated theta values
#| fig-height: 6

fig_est_theta_rain <- sim_data_g %>% 
  ggplot(aes(x = factor(time), y = est_theta, group = time)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.8) +
  geom_rain(id.long.var = "id",
            fill = "lightblue", alpha = 0.7) +
  scale_y_continuous(limits = c(-5,5), breaks = seq(-5,5,1)) +
  theme_rise()

fig_input_theta_rain <- sim_data_g %>% 
  ggplot(aes(x = factor(time), y = input_theta, group = time)) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.8) +
  geom_rain(id.long.var = "id",
            fill = "lightpink", alpha = 0.7) +
  scale_y_continuous(limits = c(-5,5), breaks = seq(-5,5,1)) +
  theme_rise()

fig_input_theta_rain / fig_est_theta_rain +
  plot_layout(axes = "collect_x")
```

### Models

We'll first make some models without measurement error, both with `lmer()`
and `brm()`. First with the input parameters to get a reference model for both.

I learnt about measurement error and brms in this excellent online book by [Solomon Kurz](https://solomonkurz.netlify.app/):
<https://bookdown.org/content/4857/missing-data-and-other-opportunities.html#measurement-error>

There are three different ways to specify measurement error, `se()`, `mi()`, and `me()`, where the last one seems to only be relevant for when you have measurement error on both predictors and outcome. Thus, we will use the two others.

```{r}

# reference model with input thetas
lm_fit0 <- lmer(input_theta ~ time + (1 | id),
              data = sim_data_g)

# estimated thetas
lm_fit1 <- lmer(est_theta ~ time + (1 | id),
              data = sim_data_g)

```

```{r}
#| output: false
# put the data into a `list()`
d <- list(
  input_theta = sim_data_g$input_theta,
  est_theta = sim_data_g$est_theta,
  est_sem = sim_data_g$est_sem,
  time = sim_data_g$time,
  id = sim_data_g$id
)

# Input/simulated thetas
b0_fit <- 
  brm(data = d, 
      family = gaussian,
      input_theta ~ time + (1 | id),
      prior = c(prior(normal(0, 3), class = Intercept),
                prior(normal(0, 2), class = b),
                prior(exponential(1), class = sigma),
                prior(exponential(1 / 0.463), class = sd)
                ),
      iter = 2000, warmup = 1000, 
      cores = 4, chains = 4,
      seed = 15)

# With estimated thetas

b1_fit <- 
  brm(data = d, 
      family = gaussian,
      est_theta ~ time + (1 | id),
      prior = c(prior(normal(0, 3), class = Intercept),
                prior(normal(0, 2), class = b),
                prior(exponential(1), class = sigma),
                prior(exponential(1 / 0.463), class = sd)
                ),
      iter = 2000, warmup = 1000, 
      cores = 4, chains = 4,
      seed = 15)


# With measurement error using se()
b2_fit <- 
  brm(data = d, 
      family = gaussian,
      est_theta | se(est_sem, sigma = TRUE) ~ time + (1 | id),
      prior = c(prior(normal(0, 3), class = Intercept),
                prior(normal(0, 2), class = b),
                prior(exponential(1), class = sigma),
                prior(exponential(1 / 0.463), class = sd)
                ),
      iter = 2000, warmup = 1000, 
      cores = 4, chains = 4,
      seed = 15)

# using mi()

b3_fit <- 
  brm(data = d, 
      family = gaussian,
      est_theta | mi(est_sem) ~ time + (1 | id),
      prior = c(prior(normal(0, 3), class = Intercept),
                prior(normal(0, 2), class = b),
                prior(exponential(1), class = sigma),
                prior(exponential(1 / 0.463), class = sd)
                ),
      iter = 2000, warmup = 1000, 
      cores = 4, chains = 4,
      seed = 15)
```

### Results

`lmer()` models.

```{r}
#| label: tbl-lmresults
#| tbl-cap: Results from lmer() models
modelsummary(list("Reference model" = lm_fit0,
                  "Estimated thetas" = lm_fit1))
```


```{r}
#| label: fig-lmresults
#| fig-cap: Results from lmer() models
modelplot(list("Reference model" = lm_fit0,
               "Estimated thetas" = lm_fit1))
```

`brm()` models.
```{r}
#| label: tbl-brmsresults
#| tbl-cap: Results from brm() models

modelsummary(list("Reference model" = b0_fit,
                  "Estimated thetas" = b1_fit,
                  "With se()" = b2_fit,
                  "With mi()" = b3_fit))
```

```{r}
#| label: fig-brmsresults
#| fig-cap: Results from brm() models
modelplot(list("Reference model" = b0_fit,
                  "Estimated thetas" = b1_fit,
                  "With se()" = b2_fit,
                  "With mi()" = b3_fit))
```


## Software used

```{r}
library(grateful)
pkgs <- cite_packages(cite.tidyverse = TRUE, 
                      output = "table",
                      bib.file = "grateful-refs.bib",
                      include.RStudio = TRUE,
                      out.dir = getwd())

formattable(pkgs,
            table.attr = 'class=\"table table-striped\" style="font-size: 15px; font-family: Lato; width: 80%"')

```

```{r}
sessionInfo()
```

## References
